\bf Effects without a Cause: The Search for Demographic Anomalies

### 2020-01-10 MANUSCRIPT/manuscript.Rmd
More edits/updates.

abstract: |
  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies. Here, we investigate US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?. We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague? totalling 164k protective deaths. 12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births. These results suggest the widespread detection of demographic anomalies. Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.

@@

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
```

@@

@@

Causal inference approaches in social sciences have a long history [@grimmer2015ppsp]. Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment. For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph]. Using natural experiments, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use]. Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers.

Outliers are then reported using a simple t-statistic.

@@

## Mortality

We begin with mortality for New York State (Figure 1). We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated. Clearly, this mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.

>>>>

abstract: |
  The proliferation of data, modern computing advances, and powerful statistical algorithms unlock the potential search for hidden, previously un- or understudied demographic anomalies. Here, we demonstrate how demographers can use causal inference techniques to identify anamolies by investigating US state-level fertility and mortality time series since 1999 to uncover the hidden baby booms/busts and mortality plagues?/non-plauges?. We find 38 states exhibited at least one mortality anomaly, totalling more than 318k anomalous deaths and an additional 6 states exhibited at least one mortality non-plague? totalling 164k protective deaths. 12 states exhibited baby busts, totalling more than 240k missing births and an additional 11 states exhibited baby booms totalling more than 134k additional births. These results suggest the widespread detection of demographic anomalies. Our analysis does not examine the *causes* of these anomalies and our results point to important further research on the causes of anomalous demographic behavior.

@@

library(knitr)
read_chunk('../R/SCRIPTS/000-Libraries.R')
read_chunk('../R/SCRIPTS/001-DataLoad.R')
```

@@

```{r DataLoad, include=FALSE}
```

@@

Causal inference, in its most common usage, is an attempt to uncover the underlying mechanism that results in changes in a phenomenon and have a long history approaches in social sciences [@grimmer2015ppsp]. Often, the identification of a casual mechanism requires either a randomized control trial (RCT) or expert knowledge applied to a natural experiment. For population-level analysis, expensive RCTs usually preclude their widespread adoption [@west2008ajph]. Using natural experiments provide less certainty about causation and fewer opportunities to replicate work but they provide important and widespread insights such as the determinants of migration, however, provides important and more widespread insights on the determinants of migration after Hurricane Katrina in 2006 [@fussellRecoveryMigrationCity2014;@horiDisplacementDynamicsSouthern2009], mini baby booms after electrical blackouts [@fetzer2018jpe], and the highly publicized estimates of excess mortality after Hurricane Maria in Puerto Rico in 2016 [@kishore2018mortality;@santos2018use]. Both RCTs and natural experiments follow traditional, hypothesis testing, inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed.

Both RCTs and natural experiments follow traditional hypothesis testing--inductive scientific paradigms where a question is first posed and scientists generate or find data to answer the question. Such approaches, while extensively utilized and time-tested, are likely to miss important phenomena that might go unnoticed. Data scientists are rethinking these approaches and have re-envisioned causal inference as a machine learning technique that allows big data researchers to move from correlations that have been established with good accuracy to making strong conclusions about the underlying mechanisms that produce these correlations
[@pearl2018book].

@@

where $\pi(B)=\sum_{i=o}^{inf} \pi_iB^i$. The identification of outliers then involves a three step process to (1) identify all potential outliers, $t_j$ and $L_j (B)$, (2) joint estimates of model parameters and outlier effects are computed to identify potentially spurious outliers, and (3) the outliers and effects are re-estimated without spurious outliers. Of importance here is step 2, which relies on some critical value above which an outlier at time point m is considered spurious. Based on Chien and Liu’s [-@chen1993joint] recommendation, we set a critical value of 3.5 which generally minimizes the possibility of Type I errors or false positive outliers. Outliers are reported using a simple t-statistic.

@@

## Mortality 
We begin with mortality for New York State (\autoref{fig:mortnewyork}). We identify seven anomalies in the mortality time series for New York, all with t-statistics in excess of 3.91, making these anomalies highly significant. The algorithm correctly identifies September 2001 as an additive outlier (2001:09 t=6.40) where mortality in that month was 1,628 higher than anticipated. This mortality event is likely caused by the September 11 tragedy and the detection of this mortality event provides confidence in our detection of other anomalies.

```{r MortalityNewYork, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New York state mortality, 1999-2016.} The top panel contains the original time series (light gray) and the corrected, counter-factual time series in the absence of anomalies (blue). The red dots on the top panel correspond to the onset of detected anomalies. The bottom panel contains the magnitude and type of the outlier. We detect all three types of outliers in New York mortality data during period 1999-2016. We detect additive outliers (AO) in January 2000, September 2001, January 2005, and January 2013; temporary change (TC) outliers in February 2007 and January 2015; and a level shift (LS) starting in February 2004. Outliers range in significance from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic of 6.40 in the AO outlier in 2001:09. \\label{fig:mortnewyork}")}

# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
dat2 <- filter(dat_mort, location == "X36")
dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)

plot(outlier.county,
     cex.axis=0.8) +
title(main = "New York - Monthly Mortality",
      adj = 1)
```

In **\autoref{fig:mortnewyork}**, notice the strong level shift (LS) that occurs in February 2004 (2004:02 t=5.59) which prevented almost 870 deaths per month. This shift totals more than 144,000 averted deaths compared to the counter-factual time series and is the single largest mortality protective anomaly among all states. This translates to 4.5% fewer deaths than expected over the time period. What is driving this mortality protection? What policies did NY put into place that might have contributed to this considerable mortality reduction? What environmental conditions may have changed? These are the kinds of questions that arise from our analyses.

Contrast the mortality protection in New York with the enhanced mortality in New Hampshire (**\autoref{fig:mortnewhamp}**). Here we detect two highly significant level shifts (LS) in the monthly mortality data, first in April 2010 and again in November 2014 (2010:04 t=3.51; 2014:11 t=6.68). These anomalies suggest New Hampshire experienced 9,700 more deaths (+14% more than expected) in a seven-year period beginning in early 2010. This is the single largest percentage mortality increase/decrease we detected among all states. Not coincidentally, NH has the second highest opioid-related mortality in the US [@beetham2019access] and it is reasonable that we detect this epidemic in our results.

```{r MortalityNewHamp, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("\\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.} Here we detect two outliers, both level shifts in outliers (LS) in April 2010 and again in November 2014. These anomalies suggest New Hampshire experienced approximately 9,700 more deaths than expected since 2010 or 14% more deaths in the state over just seven years. \\label{fig:mortnewhamp}")}

# df2<- setNames(data.frame(matrix(ncol = 5, nrow = 0)), c("type", "ind", "time", "coefhat", "tstat"))
dat2 <- filter(dat_mort, location == "X33")
dat3 <- ts(dat2$Target, start = year(min(dat_mort$Year)), end = c(year(max(dat_mort$Year)),12), frequency = 12)
outlier.county <- tsoutliers::tso(dat3,types = c("AO","LS","TC"),cval = sigma, maxit.iloop=10)

plot(outlier.county,
     cex.axis=0.8) +
title(main = "New Hampshire - Monthly Mortality",
      adj = 1)
```

In the case of New York and New Hampshire, some of the demographic anomalies have plausible explanations. It seems likely that New York’s AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise in New Hampshire’s mortality starting in 2010 could be linked to the opioid epidemic. However, we detect numerous seemingly unexplainable demographic anomalies in other states. Figure 3 shows two such unexplained anomalies.

### 2020-01-10 MANUSCRIPT/manuscript.tex
More edits/updates.

\bigskip
\begin{abstract}
The proliferation of data, modern computing advances, and powerful
statistical algorithms unlock the potential search for hidden,
previously un- or understudied demographic anomalies. Here, we
investigate US state-level fertility and mortality time series since
1999 to uncover the hidden baby booms/busts and mortality
plagues?/non-plauges?. We find 38 states exhibited at least one
mortality anomaly, totalling more than 318k anomalous deaths and an
additional 6 states exhibited at least one mortality non-plague?
totalling 164k protective deaths. 12 states exhibited baby busts,
totalling more than 240k missing births and an additional 11 states
exhibited baby booms totalling more than 134k additional births. These
results suggest the widespread detection of demographic anomalies. Our
analysis does not examine the \emph{causes} of these anomalies and our
results point to important further research on the causes of anomalous
demographic behavior.\\
\end{abstract}

@@

Causal inference approaches in social sciences have a long history
\citep{grimmer2015ppsp}. Often, the identification of a casual mechanism
requires either a randomized control trial (RCT) or expert knowledge
applied to a natural experiment. For population-level analysis,
expensive RCTs usually preclude their widespread adoption
\citep{west2008ajph}. Using natural experiments, however, provides
important and more widespread insights on the determinants of migration
after Hurricane Katrina in 2006
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
the highly publicized estimates of excess mortality after Hurricane
Maria in Puerto Rico in 2016
\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
experiments follow traditional, hypothesis testing, inductive scientific
paradigms where a question is first posed and scientists generate or
find data to answer the question. Such approaches, while extensively
utilized and time-tested, are likely to miss important phenomena that
might go unnoticed.

@@

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chien and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers.

Outliers are then reported using a simple t-statistic.

@@

We detect numerous anomalous mortality and fertility events at the US
state-level since 1999. A full listing of these anomalies can be found
in the \textbf{Supplementary Materials}. Here, we highlight highly
significant mortality and fertility events across all three types of
outliers (Additive Outliers, Level Shift Outliers, and Temporary Change
Outliers) where we have identified plausible explanations for these
outliers.

@@

We begin with mortality for New York State (Figure 1). We identify seven
anomalies in the mortality time series for New York, all with
t-statistics in excess of 3.91, making these anomalies highly
significant. The algorithm correctly identifies September 2001 as an
additive outlier (2001:09 t=6.40) where mortality in that month was
1,628 higher than anticipated. Clearly, this mortality event is likely
caused by the September 11 tragedy and the detection of this mortality
event provides confidence in our detection of other anomalies.

>>>>

\bigskip
\begin{abstract}
The proliferation of data, modern computing advances, and powerful
statistical algorithms unlock the potential search for hidden,
previously un- or understudied demographic anomalies. Here, we
demonstrate how demographers can use causal inference techniques to
identify anamolies by investigating US state-level fertility and
mortality time series since 1999 to uncover the hidden baby booms/busts
and mortality plagues?/non-plauges?. We find 38 states exhibited at
least one mortality anomaly, totalling more than 318k anomalous deaths
and an additional 6 states exhibited at least one mortality non-plague?
totalling 164k protective deaths. 12 states exhibited baby busts,
totalling more than 240k missing births and an additional 11 states
exhibited baby booms totalling more than 134k additional births. These
results suggest the widespread detection of demographic anomalies. Our
analysis does not examine the \emph{causes} of these anomalies and our
results point to important further research on the causes of anomalous
demographic behavior.\\
\end{abstract}

@@

Causal inference, in its most common usage, is an attempt to uncover the
underlying mechanism that results in changes in a phenomenon and have a
long history approaches in social sciences \citep{grimmer2015ppsp}.
Often, the identification of a casual mechanism requires either a
randomized control trial (RCT) or expert knowledge applied to a natural
experiment. For population-level analysis, expensive RCTs usually
preclude their widespread adoption \citep{west2008ajph}. Using natural
experiments provide less certainty about causation and fewer
opportunities to replicate work but they provide important and
widespread insights such as the determinants of migration, however,
provides important and more widespread insights on the determinants of
migration after Hurricane Katrina in 2006
\citep{fussellRecoveryMigrationCity2014, horiDisplacementDynamicsSouthern2009},
mini baby booms after electrical blackouts \citep{fetzer2018jpe}, and
the highly publicized estimates of excess mortality after Hurricane
Maria in Puerto Rico in 2016
\citep{kishore2018mortality, santos2018use}. Both RCTs and natural
experiments follow traditional, hypothesis testing, inductive scientific
paradigms where a question is first posed and scientists generate or
find data to answer the question. Such approaches, while extensively
utilized and time-tested, are likely to miss important phenomena that
might go unnoticed.

Both RCTs and natural experiments follow traditional hypothesis
testing--inductive scientific paradigms where a question is first posed
and scientists generate or find data to answer the question. Such
approaches, while extensively utilized and time-tested, are likely to
miss important phenomena that might go unnoticed. Data scientists are
rethinking these approaches and have re-envisioned causal inference as a
machine learning technique that allows big data researchers to move from
correlations that have been established with good accuracy to making
strong conclusions about the underlying mechanisms that produce these
correlations \citep{pearl2018book}.

@@

where \(\pi(B)=\sum_{i=o}^{inf} \pi_iB^i\). The identification of
outliers then involves a three step process to (1) identify all
potential outliers, \(t_j\) and \(L_j (B)\), (2) joint estimates of
model parameters and outlier effects are computed to identify
potentially spurious outliers, and (3) the outliers and effects are
re-estimated without spurious outliers. Of importance here is step 2,
which relies on some critical value above which an outlier at time point
m is considered spurious. Based on Chien and Liu's
\citeyearpar{chen1993joint} recommendation, we set a critical value of
3.5 which generally minimizes the possibility of Type I errors or false
positive outliers. Outliers are reported using a simple t-statistic.

@@

We detect numerous anomalous mortality and fertility events at the US
state-level since 1999. A full listing of these anomalies can be found
in the \textbf{Supplementary Materials}. We begin with a summary of the
anomalies we detect and then we highlight highly significant mortality
and fertility anomalies across all three types of outliers (Additive
Outliers, Level Shift Outliers, and Temporary Change Outliers) with
plausible explanations.

\hypertarget{overall-anomalies}{%
\subsection{Overall Anomalies}\label{overall-anomalies}}

@@

We begin with mortality for New York State (\autoref{fig:mortnewyork}).
We identify seven anomalies in the mortality time series for New York,
all with t-statistics in excess of 3.91, making these anomalies highly
significant. The algorithm correctly identifies September 2001 as an
additive outlier (2001:09 t=6.40) where mortality in that month was
1,628 higher than anticipated. This mortality event is likely caused by
the September 11 tragedy and the detection of this mortality event
provides confidence in our detection of other anomalies.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/MortalityNewYork-1.pdf}
\caption{\textbf{Anomaly Detection for New York state mortality, 1999-2016.}
The top panel contains the original time series (light gray) and the
corrected, counter-factual time series in the absence of anomalies
(blue). The red dots on the top panel correspond to the onset of
detected anomalies. The bottom panel contains the magnitude and type of
the outlier. We detect all three types of outliers in New York mortality
data during period 1999-2016. We detect additive outliers (AO) in
January 2000, September 2001, January 2005, and January 2013; temporary
change (TC) outliers in February 2007 and January 2015; and a level
shift (LS) starting in February 2004. Outliers range in significance
from a low t-statistic of 3.91 in the TC 2015:01 to a high t-statistic
of 6.40 in the AO outlier in 2001:09. \label{fig:mortnewyork}}
\end{figure}

\begin{verbatim}
## numeric(0)
\end{verbatim}

In \textbf{\autoref{fig:mortnewyork}}, notice the strong level shift
(LS) that occurs in February 2004 (2004:02 t=5.59) which prevented
almost 870 deaths per month. This shift totals more than 144,000 averted
deaths compared to the counter-factual time series and is the single
largest mortality protective anomaly among all states. This translates
to 4.5\% fewer deaths than expected over the time period. What is
driving this mortality protection? What policies did NY put into place
that might have contributed to this considerable mortality reduction?
What environmental conditions may have changed? These are the kinds of
questions that arise from our analyses.

Contrast the mortality protection in New York with the enhanced
mortality in New Hampshire (\textbf{\autoref{fig:mortnewhamp}}). Here we
detect two highly significant level shifts (LS) in the monthly mortality
data, first in April 2010 and again in November 2014 (2010:04 t=3.51;
2014:11 t=6.68). These anomalies suggest New Hampshire experienced 9,700
more deaths (+14\% more than expected) in a seven-year period beginning
in early 2010. This is the single largest percentage mortality
increase/decrease we detected among all states. Not coincidentally, NH
has the second highest opioid-related mortality in the US
\citep{beetham2019access} and it is reasonable that we detect this
epidemic in our results.

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/MortalityNewHamp-1.pdf}
\caption{\textbf{Anomaly Detection for New Hampshire state mortality, 1999-2016.}
Here we detect two outliers, both level shifts in outliers (LS) in April
2010 and again in November 2014. These anomalies suggest New Hampshire
experienced approximately 9,700 more deaths than expected since 2010 or
14\% more deaths in the state over just seven years.
\label{fig:mortnewhamp}}
\end{figure}

\begin{verbatim}
## numeric(0)
\end{verbatim}

In the case of New York and New Hampshire, some of the demographic
anomalies have plausible explanations. It seems likely that New York's
AO anomaly in September 2001 is caused by the 9/11 tragedy and the rise
in New Hampshire's mortality starting in 2010 could be linked to the
opioid epidemic. However, we detect numerous seemingly unexplainable
demographic anomalies in other states. Figure 3 shows two such
unexplained anomalies.
